A Large-scale Study of Polymorphism in R

### 2019-04-02 paper/main.tex
changes all over

@@

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|l@{}}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}  & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double} & vector of 64 bit floating points\\
\k{complex} & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw} & vector of bytes\\
\k{list} & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}  &  singleton null value\\
\k{S4}    &  instance of a S4 class \\
\k{closure} & a function with its environment\\
\k{environment} & a mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
\multicolumn{2}{l}{\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise}}\\
\multicolumn{2}{l}{
\k{language},
\k{char},
\k{...}, 
\k{any},
\k{expression},
}\\
\multicolumn{2}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

@@

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L                     
[1] 3
> x <- c(1,2,3,4)            
> attr(x,"dim") <- c(2,2)    
> x + c(1,2)                 
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} & 
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+` &\tt D[1], I[1] -> D[1] \\
\tt c& D[1], D[1], D[1], D[1] -> D[4]\\
\tt `<-` &\tt S, D[3] -> D[3]\\
\tt c & \tt D[1], D[1] -> D[2]\\
\tt `attr<-1`&\tt  S, C, D[2] -> \\
\tt c &\tt D[1], D[1] -> D[2]\\
\tt `+` &\tt D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

@@

\section{Stuff}

@@

\begin{itemize}
  \item an argument is {\it polymorphic in type} if it has been called with at
  least two distinct types according to our definition of type.

   \item an argument is {\it polymorphic in class} if it has been called with at least two distinct classes according to the \code{class} function.

  \item attributes are a bit more tricky.  We can't simply count the number of
  attributes that have inhabited an argument, since values can contain
  arbitrarily many attributes (and having several attributes isn't
  polymorphism---think records).  Instead, we define the {\it attribute
    pattern} of a value to be the set of all attribute name and type pairs
  in the attribute list of a value (obtained via \code{attributes}).  And so
  an argument is {\it polymorphic in attribute} if it has been called with
  at least two distinct attribute patterns.

\end{itemize}

@@

\begin{itemize}

\item data frames have three attributes: {\tt names} for column names, {\tt
  row.names} for row names, and {\tt class} (which is instantiated to {\tt
  "data.frame"} upon creation).

\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names. Curiously, the {\tt dim} attribute is all that is needed to define a matrix (i.e., a list with a {\tt dim} attribute {\it is} a matrix).

\item lists and vectors may have a {\tt names} attribute, which assigns
  names to locations.

\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \texttt{"factor"}.

\end{itemize}

@@

%
%
%
%
\subsection{Granularity of Type Information}

@@

\begin{itemize}

	\item if the analysis encounters \code{ typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.
	This is also true for integer, complex, raw, logical, and character {\tt typeof}s (i.e., the primitive vectorized data types).
	A notable imperfection here is is that unit length vectors will appear as scalars, so later in our analysis we ensure that argument signatures containing both scalar/X and vector/X are collapsed down to contain only vector/X.

@@

	\item similarly, if it encounters a data frame (i.e., class {\tt data.frame} with appropriate attributes), we ascribe the type {\tt data.frame}.

\end{itemize}

%
%
%
%
%
%
\section{Examples of Polymorphism}
\label{sec:polyex}

\AT{Putting examples of polymorphism in this section, we can talk about them and where they should go.}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
    ...
    if (is.character(fn)) {
        if (fn == "MODE" | fn == "MAP") {
            do_MAP <- TRUE
        }
        else {
            do_MAP <- FALSE
            fn <- match.fun(fn) # looks up the function name
        }
    }
    ...
\end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {
  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat)) {
        stop(simpleError("...'dat' must be a vector!"))
      }
    ...
  } else {
    ...
    # here, dat is data.frame
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
fixGenes <- function (network, fixIndices, values) {
  ...
  network$fixed[fixIndices] <- as.integer(values)
  ...
}
\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
tee <- function (x, theta, D1, D2, phi)  {
    if (is.vector(x)) {
        x <- t(x)
    }
    ...
\end{lstlisting}}\caption{Example of matrix/(\dblv) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}

%
%
%
%
%
%
\section{Results}\label{sec:results}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:13:17 2019
\begin{table}[ht]
\label{tab:L0toppoly}
\centering
\begin{tabular}{lrr}
  \hline
type & count & perc \\ 
  \hline
  double, integer & 19,799 & 28.73 \\ 
  double, list & 5,324 & 7.72 \\ 
  double, NULL & 5,152 & 7.48 \\ 
  character, NULL & 3,682 & 5.34 \\ 
  double, logical & 3,388 & 4.92 \\ 
  character, double & 2,842 & 4.12 \\ 
  list, NULL & 2,799 & 4.06 \\ 
  character, list & 1,836 & 2.66 \\ 
  character, integer & 1,663 & 2.41 \\ 
  character, double, integer & 1,362 & 1.98 \\ 
   \hline
\end{tabular}
\caption{Top polymorphic signatures for the L0 type system.}
\end{table}

@@

	\item \{double, integer\}: in R, doubles and integers are effectively interchangeable, as casting from one to the other is straightforward and automatically performed when appropriate.
	For example, \code{c(1, 2)[1.9] == 1}.
	This distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.

	\item \{double, list\} and \{character, list\}: at first glance, these type signatures are suspect.
	Why would so many programmers want to pass a double or a list to some function?
	The key insight here is that doubles and characters are {\it vectorized} in R, meaning that a double is really a vector of doubles.
	Thus, these seemingly strange signatures are revealed to be entirely reasonable: these signatures are likely to describe arguments which are either vectors or lists.

	\item \{X, NULL\}: in most languages, NULL is an inhabitant of all types, and we see that this is not the case in R.
	That said, there is a pattern of many functions having default or optional arguments, and in many of these cases NULL is the default value.

	\item \{double, logical\}: the first values of type logical that come to mind are TRUE and FALSE, which does little to explain this strange signature.
	Indeed, the key here is an implementation detail of R's: recall that NAs inhabit every type.
	As it happens, the default type of NA is logical (e.g., \code{typeof(NA) == "logical"}), and this is likely what is being seen here.
	We will control for this by distinguishing NAs with their own type, {\tt raw\_NA}, but we will get to that in time.

	\item \{character, double\}, \{character, integer\} and \{character, double, integer\}: yet more strange signatures.
	These signatures are a little more subtle to explain.
	For one, indexing lists and vectors can be done with either strings (i.e., to access a named index) or with numbers.
	Another option is dates: R has many ways to present dates, and strings and numbers feature among them.
	Finally, many \AT{get numbers} functions in R return a string if an error occurs during execution.

>>>>

\newcommand{\attr}[2]{\ensuremath{#1\,({\mathit #2})}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % Any

@@

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|c|l@{}}\hline
\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & \L & vector of boolean values\\
\k{integer}   & \I & vector of 32 bit integer values\\
\k{double}    & \D & vector of 64 bit floating points\\
\k{complex}   & \X & vector of complex values\\
\k{character} & \C & vector of strings values\\
\k{raw}       & \R & vector of bytes\\
\k{list}      & \l & vector of values of any type\\\hline
\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      & \sN &  singleton null value\\
\k{S4}        & \sS &  instance of a S4 class \\
\k{closure}   & \sF & a function with its environment\\
\k{environment}&\sE &  a mapping from symbol to value \\\hline
\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
\multicolumn{3}{l}{\k{special},
\k{builtin},
\k{symbol} (\sY),
\k{pairlist},
\k{promise}}\\
\multicolumn{3}{l}{
\k{language},
\k{char},
\k{...}, 
\k{any},
\k{expression},
}\\
\multicolumn{3}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

@@

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L                     
[1] 3
> x <- c(1,2,3,4)            
> attr(x,"dim") <- c(2,2)    
> x + c(1,2)                 
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} & 
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+`: &\tt \sD \sI \to \sD \\
\tt c:& \sD \sD \sD \sD \to \D\\
\tt `<-`: &\tt \sY \D \to \D\\
\tt c: & \tt \sD \sD \to \D\\
\tt `attr<-`:&\tt  \sY \C \D \to \\
\tt c: &\tt \sD \sD \to \D\\
\tt `+`: &\tt \attr\D{dim=\D} \D \to \attr\D{dim=\D}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

@@

\section{Types for R}

@@

\begin{itemize}
\item an argument is {\it polymorphic in type} if it has been called with at
  least two distinct types according to our definition of type.
\item an argument is {\it polymorphic in class} if it has been called with
  at least two distinct classes according to the \code{class} function.
\item attributes are a bit more tricky.  We can't simply count the number of
  attributes that have inhabited an argument, since values can contain
  arbitrarily many attributes (and having several attributes isn't
  polymorphism---think records).  Instead, we define the {\it attribute
    pattern} of a value to be the set of all attribute name and type pairs
  in the attribute list of a value (obtained via \code{attributes}).  And so
  an argument is {\it polymorphic in attribute} if it has been called with
  at least two distinct attribute patterns.
\end{itemize}

@@

\begin{itemize}
\item data frames have three attributes: {\tt names} for column names, {\tt
  row.names} for row names, and {\tt class} (which is instantiated to {\tt
  "data.frame"} upon creation).
\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
  list with a {\tt dim} attribute {\it is} a matrix).
\item lists and vectors may have a {\tt names} attribute, which assigns
  names to locations.
\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \texttt{"factor"}.
\end{itemize}

@@

\subsection{Granularity of Type Information}

@@

\begin{itemize}
\item if the analysis encounters \code{ typeof(x) == "double"}, it will look
  at the length of {\tt x} to determine if {\tt x} is a scalar or a vector,
  generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.  This is
  also true for integer, complex, raw, logical, and character {\tt typeof}s
  (i.e., the primitive vectorized data types).  A notable imperfection here
  is is that unit length vectors will appear as scalars, so later in our
  analysis we ensure that argument signatures containing both scalar/X and
  vector/X are collapsed down to contain only vector/X.

@@

\item similarly, if it encounters a data frame (i.e., class {\tt data.frame}
  with appropriate attributes), we ascribe the type {\tt data.frame}.
\end{itemize}

%
%
%
\section{Examples of Polymorphism}
\label{sec:polyex}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
    ...
    if (is.character(fn)) 
        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
        else {
            do_MAP <- FALSE
            fn <- match.fun(fn) # looks up the function name
        }
\end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {
  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
    ...
  } else {
    ...
    # here, dat is data.frame
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
fixGenes <- function (network, fixIndices, values) {
  ...
  network$fixed[fixIndices] <- as.integer(values)
  ...
\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
tee <- function (x, theta, D1, D2, phi)  {
    if (is.vector(x)) x <- t(x)
    ...
\end{lstlisting}}\caption{Example of matrix/(\dblv) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}

%
%
%
\section{Results}\label{sec:results}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:13:17 2019
\begin{table}[ht]\label{tab:L0toppoly}\centering
\begin{tabular}{lrr}  \hline
type & count & perc \\\hline\\[-3mm] 
  \D, \I & 19,799 & 28.73 \\[1mm] 
  \D, \l & 5,324 & 7.72 \\ [1mm] 
  \D, \sN & 5,152 & 7.48 \\ [1mm] 
  \C, \sN & 3,682 & 5.34 \\ [1mm] 
  \D, \L & 3,388 & 4.92 \\ [1mm] 
  \C, \D & 2,842 & 4.12 \\ [1mm] 
  \l, \sN & 2,799 & 4.06 \\ [1mm] 
  \C, \l & 1,836 & 2.66 \\ [1mm] 
  \C, \I & 1,663 & 2.41 \\ [1mm] 
  \C, \I, \D & 1,362 & 1.98 \\[1mm]   \hline
\end{tabular}
\caption{Top polymorphic signatures for the L0 type system.}
\end{table}

@@

\begin{enumerate}
\item \{\D, \I\}: in R, doubles and integers are effectively
  interchangeable, as casting from one to the other is straightforward and
  automatically performed when appropriate.  For example, \code{c(1, 2)[1.9]
    == 1}.  This distinction is mainly just relevant to the implementation,
  as integers and doubles are stored and dealt with differently in the
  runtime but programmers only see the distinction when printing.

\item \{\D, \l\} and \{\C, \l\}: at first glance, these type signatures are
  suspect.  Why would so many programmers want to pass a double or a list to
  some function?  The key insight here is that doubles and characters are
  {\it vectorized} in R, meaning that a double is really a vector of
  doubles.  Thus, these seemingly strange signatures are revealed to be
  entirely reasonable: these signatures are likely to describe arguments
  which are either vectors or lists.

\item \{\ANY,\sN\}: in most languages, NULL is an inhabitant of all types,
  and we see that this is not the case in R.  That said, there is a pattern
  of many functions having default or optional arguments, and in many of
  these cases NULL is the default value.

\item \{\D,\L\}: the first values of type logical that come to mind are TRUE
  and FALSE, which does little to explain this strange signature.  Indeed,
  the key here is an implementation detail of R's: recall that NAs inhabit
  every type.  As it happens, the default type of NA is logical (e.g.,
  \code{typeof(NA) == "logical"}), and this is likely what is being seen
  here.  We will control for this by distinguishing NAs with their own type,
  {\tt raw\_NA}, but we will get to that in time.

\item \{\C,\D\}, \{\C,\I\} and \{character, double, integer\}: yet more
  strange signatures.  These signatures are a little more subtle to explain.
  For one, indexing lists and vectors can be done with either strings (i.e.,
  to access a named index) or with numbers.  Another option is dates: R has
  many ways to present dates, and strings and numbers feature among them.
  Finally, many \AT{get numbers} functions in R return a string if an error
  occurs during execution.
